---
title: "Software and Code, and Data Availability"
author: "Robert M Flight"
date: last-modified
date-format: YYYY-MM-DD HH:mm
format: docx
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE)
source(here::here("packages.R"))
```

```{r}
#| label: load-targets
#| include: false
tar_load(dependencies)

get_package_version = function(dependency_obj, package_name) {
  #dependency_obj = dependencies
  #package_name = "ComplexHeatmap"
  tmp_df = dependency_obj$package |>
    dplyr::filter(package %in% package_name)
  if (is.na(tmp_df$loadedversion)) {
    use_version = tmp_df$ondiskversion
  } else {
    use_version = tmp_df$loadedversion
  }
  package_version = paste0(tmp_df$package, " v", use_version)
  if (grepl("Github", tmp_df$source)) {
    package_version = paste0(package_version, " (", gsub("\\(", "", strtrim(tmp_df$source, nchar(tmp_df$source) - 34)), ")")
  }
  package_version
}
```

## Software, Code and Data Availability

All calculations were performed using `r dependencies$platform$version`, with the following packages and versions: 

* creation of raincloud plots: `{r} get_package_version(dependencies, "ggrain")`
* general plotting: `{r} get_package_version(dependencies, "ggplot2")`
* calculation of principal component variances and associates: `{r} get_package_version(dependencies, "visualizationQualityControl")`
* receiver operating curves and plots: `{r} get_package_version(dependencies, "pROC")`
* UpSet plots and heatmaps: `{r} get_package_version(dependencies, "ComplexHeatmap")`
* ICI-Kendall-tau calculations: `{r} get_package_version(dependencies, "ICIKendallTau")`
* Hierarchical clustering reordering: `{r} get_package_version(dependencies, "dendsort")`
* Workflow control: `{r} get_package_version(dependencies, "targets")`

A full list of all installed R packages can be found in the renv.lock file in the Zenodo repository at https://dx.doi.org/10.5281/zenodo.12190086.

All code used for the analysis is available on GitHub at https://github.com/MoseleyBioinformaticsLab/conwell-manuscript.ChronicPancreatitisUrineMetabolomics/, and on Zenodo at https://dx.doi.org/10.5281/zenodo.12190086.

